---
title: "Transfer Portal Models v1"
author: "Gregory Lederer"
date: "2025-07-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = "/Users/gregorylederer/Desktop/Transfer Portal Model")
```

#packages

```{r}
library(httr2)
library(jsonlite)
library(data.table)
library(dplyr)
library(purrr)
library(hoopR)
library(cbbdata)
library(tidyverse)
library(stringr)
library(readr)
library(forcats)
library(broom)
library(pROC)
library(caret)
library(car)
library(ggplot2)
library(boot)
library(scales)
library(nnet)
library(glmnet)
library(lubridate)
library(tibble)
```

2026 player data

```{r}
url <- "https://www.barttorvik.com/getadvstats.php?year=2026"

txt <- request(url) |>
  req_user_agent("cbb-analytics (personal use)") |>
  req_perform() |>
  resp_body_string()

rows <- fromJSON(txt, simplifyVector = FALSE)

df <- rbindlist(lapply(rows, as.list), fill = TRUE, use.names = FALSE)
df <- as.data.frame(df)

cn67 <- c(
  'player', 'team', 'conf', 'GP', 'MIN.', 'ORTG', 'USG', 'eFG.', 'TS.', 'ORB.', 'DRB.', 'AST.', 'TO.',
  'FTM', 'FTA', 'FT.', 'TWO_Make', 'TWO_Att', 'TWO.', 'TPM', 'TPA', 'TP.', 'BLK.', 'STL.', 'FTR',
  'class', 'height', 'num', 'PORPAG', 'DELETE3', 'DELETE4', 'year', 'pid', 'Hometown', 'DELETE1', 'AST_TOV',
  'RIM_Make', 'RIM_Att', 'MID_Make', 'MID_Att', 'Rim.', 'Mid.', 'DUNKS_Made', 'DUNKS_Att', 'DUNK.',
  'DELETE2', 'DELETE9', 'ADV_DRTG', 'D_PORPAG', 'DELETE5', 'DELETE6', 'DELETE7', 'DELETE8', 
  'BPM', 'MP', 'OBPM', 'DBPM', 'ORB', 'DRB', 'REB', 'AST', 'STL', 'BLK', 'PTS', 'Role', 'DELETE10', 'Birthdate'
)

stopifnot(ncol(df) == length(cn67))
names(df) <- cn67

df <- df %>%
  select(-starts_with("DELETE"))

player_data_2026 <- df
```

2015 - 25 player data

```{r}
cn67 <- c(
  'player', 'team', 'conf', 'GP', 'MIN.', 'ORTG', 'USG', 'eFG.', 'TS.', 'ORB.', 'DRB.', 'AST.', 'TO.',
  'FTM', 'FTA', 'FT.', 'TWO_Make', 'TWO_Att', 'TWO.', 'TPM', 'TPA', 'TP.', 'BLK.', 'STL.', 'FTR',
  'class', 'height', 'num', 'PORPAG', 'DELETE3', 'DELETE4', 'year', 'pid', 'DELETE11', 'DELETE1', 'AST_TOV',
  'RIM_Make', 'RIM_Att', 'MID_Make', 'MID_Att', 'Rim.', 'Mid.', 'DUNKS_Made', 'DUNKS_Att', 'DUNK.',
  'DELETE2', 'DELETE9', 'ADV_DRTG', 'D_PORPAG', 'DELETE5', 'DELETE6', 'DELETE7', 'DELETE8',
  'BPM', 'MP', 'OBPM', 'DBPM', 'ORB', 'DRB', 'REB', 'AST', 'STL', 'BLK', 'PTS', 'Role', 'DELETE10', 'Birthdate'
)

get_torvik_advstats <- function(season_year, cn = cn67) {
  url <- sprintf("https://www.barttorvik.com/getadvstats.php?year=%d", season_year)

  txt <- request(url) |>
    req_user_agent("cbb-analytics (personal use)") |>
    req_perform() |>
    resp_body_string()

  rows <- fromJSON(txt, simplifyVector = FALSE)

  df <- rbindlist(lapply(rows, as.list), fill = TRUE, use.names = FALSE) |>
    as.data.frame(stringsAsFactors = FALSE)

  stopifnot(ncol(df) == length(cn))
  names(df) <- cn

  df %>%
    select(-starts_with("DELETE")) %>%
    mutate(
      num = as.character(num)
    )
}

years <- 2015:2025

all_players_by_season <- map_dfr(
  years,
  ~ get_torvik_advstats(.x),
  .id = "season_id"
) %>%
  select(-season_id)
```

team data

```{r}
cbb <- map_df(years, function(y){
  read_csv(paste0("https://barttorvik.com/", y, "_team_results.csv")) %>%
    mutate(season = y) %>%
    select(1:43, season)
})

cbb_26 <- map_df(2026, function(y){
  read_csv(paste0("https://barttorvik.com/", y, "_team_results.csv")) %>%
    mutate(season = y) %>%
    select(1:43, season)
})
```

join

```{r}
players_model <- all_players_by_season %>%
  select(-matches("^Unnamed")) %>%
  mutate(year = as.integer(year), team = str_trim(team), conf = str_trim(conf))

players_model_26 <- player_data_2026 %>%
  select(-matches("^Unnamed")) %>%
  mutate(year = as.integer(year), team = str_trim(team), conf = str_trim(conf))

teams_model <- cbb %>%
  select(team, conf, season, record, `Con Rec.`, barthag, adjoe, `oe Rank`, adjde, `de Rank`, sos, ncsos, consos, `Qual Barthag`, `Qual Games`, `Conf Win%`, WAB, `WAB Rk`) %>%
  mutate(season = as.integer(season), team = str_trim(team), conf = str_trim(conf)) %>%
  separate(record, into = c("team_wins", "team_losses"), sep = "-", remove = FALSE, convert = TRUE) %>%
  separate(`Con Rec.`, into = c("conf_wins", "conf_losses"), sep = "-", remove = FALSE, convert = TRUE)

model_table_START <- players_model %>%
  left_join(teams_model, by = c("team", "conf", "year" = "season"))

#write.csv(model_table, 'Modeling Table 2015-25.csv')
#write.csv(model_table[1:10,], 'Modeling Table 2015-25 Sample.csv')

teams_model_26 <- cbb_26 %>%
  select(team, conf, season, record, `Con Rec.`, barthag, adjoe, `oe Rank`, adjde, `de Rank`, sos, ncsos, consos, `Qual Barthag`, `Qual Games`, `Conf Win%`, WAB, `WAB Rk`) %>%
  mutate(season = as.integer(season), team = str_trim(team), conf = str_trim(conf)) %>%
  separate(record, into = c("team_wins", "team_losses"), sep = "-", remove = FALSE, convert = TRUE) %>%
  separate(`Con Rec.`, into = c("conf_wins", "conf_losses"), sep = "-", remove = FALSE, convert = TRUE)

complete_data_2026 <- players_model_26 %>%
  left_join(teams_model_26, by = c("team", "conf", "year" = "season"))
```

transfer likelihood model

```{r, eval=FALSE}
model_table <- model_table_START %>%
  mutate(
    year = as.integer(year),
    pid = as.character(pid),
    player = str_squish(player),
    team = str_squish(team),

    conf = str_squish(conf),
    conf = str_to_lower(conf),
    conf = if_else(conf == "ind", "independent", conf),

    major = case_when(
      conf %in% c("b12", "acc", "b10", "sec", "be", "p12") ~ "high",
      TRUE ~ "mid"
    ),

    conf = factor(conf),
    major = factor(major, levels = c("high", "mid")),

    class = str_squish(class),
    class = na_if(class, ""),
    class = na_if(class, "N/A"),
    class = na_if(class, "--"),
    class = case_when(
      class %in% c("Fr", "So", "Jr", "Sr") ~ class,
      TRUE ~ NA_character_
    ),

    Role = str_squish(Role),
    Role = na_if(Role, ""),

    Birthdate = as.character(Birthdate),

    underclass = case_when(
      class %in% c("Fr", "So") ~ "Underclassmen",
      class %in% c("Jr", "Sr") ~ "Upperclassmen",
      TRUE ~ NA_character_
    ),
    
    player_team_gap = BPM - barthag
  ) %>%
  group_by(team, year) %>%
  mutate(
    mp_rank = min_rank(desc(MP)),
    bpm_rank = min_rank(desc(BPM)),
    usage_rank = min_rank(desc(USG))
  ) %>%
  ungroup() %>%
  group_by(conf, year) %>%
  mutate(
    bpm_pct_conf = percent_rank(BPM),
    porpag_pct_conf = percent_rank(PORPAG)
  ) %>%
  ungroup() %>%
  group_by(major, year) %>%
  mutate(
    usg_rank_c1 = min_rank(desc(USG)),
    porpag_rank_c1 = min_rank(desc(PORPAG)),
    bpm_rank_c1 = min_rank(desc(BPM))
  ) %>%
  ungroup() %>%
  mutate(
    underused_star = if_else(BPM > 5 & mp_rank > 5, 1, 0)
  ) %>%
  group_by(team, year) %>%
  mutate(
    team_total_mp = sum(MP, na.rm = TRUE),
    minutes_share = MP / team_total_mp,
    mp_pct_team = percent_rank(MP),
    bpm_pct_team = percent_rank(BPM),
    usg_pct_team = percent_rank(USG)
  ) %>%
  ungroup() %>%
  mutate(
    bpm_minutes_gap = BPM - minutes_share,
    mid_major_star = if_else(major == "mid" & bpm_pct_conf > 0.9, 1, 0),
    low_role_high_output = if_else(porpag_pct_conf >= 0.75 & mp_pct_team <= 0.40, 1L, 0L),
    mid_major_star = if_else(major == "mid" & bpm_pct_conf >= 0.90, 1L, 0L),
    high_major_low_role = if_else(major == "high" & mp_pct_team <= 0.25, 1L, 0L),

    winning_team = if_else(!is.na(team_wins) & !is.na(team_losses) & team_wins > team_losses, 1L, 0L),
    conf_winning_team = if_else(
      !is.na(conf_wins) & !is.na(conf_losses) & conf_wins > conf_losses,
      1L, 0L
    )
  )

next_year_lookup <- model_table %>%
  select(
    pid, year, player, Birthdate,
    next_team = team,
    next_conf = conf,
    next_class = class,
    next_role = Role
  ) %>%
  distinct() %>%
  mutate(year = year - 1)

transfer_table <- model_table %>%
  left_join(
    next_year_lookup,
    by = c("pid", "year", "player", "Birthdate")
  ) %>%
  mutate(
    returned_next_year = if_else(!is.na(next_team), 1L, 0L),
    transfer_next_year = case_when(
      is.na(next_team) ~ NA_integer_,
      next_team != team ~ 1L,
      next_team == team ~ 0L
    ),
    transfer_status = case_when(
      is.na(next_team) ~ "no_next_year_row",
      next_team != team ~ "transferred",
      next_team == team ~ "stayed"
    )
  )

# start modeling
# only use rows where transfer outcome is actually known
model_data <- transfer_table %>%
  filter(year <= 2024) %>%
  filter(!is.na(transfer_next_year))

train <- model_data %>% filter(year <= 2022)
valid <- model_data %>% filter(year == 2023)
test  <- model_data %>% filter(year == 2024)

# force factor levels to be based only on training data
class_levels <- sort(unique(na.omit(train$class)))
role_levels  <- sort(unique(na.omit(train$Role)))

train <- train %>%
  mutate(
    class = factor(class, levels = class_levels),
    Role = factor(Role, levels = role_levels)
  )

valid <- valid %>%
  mutate(
    class = factor(class, levels = class_levels),
    Role = factor(Role, levels = role_levels)
  )

test <- test %>%
  mutate(
    class = factor(class, levels = class_levels),
    Role = factor(Role, levels = role_levels)
  )

# drop rows with missing predictors
train_model <- train %>%
  filter(
    complete.cases(
      transfer_next_year, class, MP, BPM, USG, ORTG,
      team_wins, barthag, `Conf Win%`, Role
    )
  )

valid_model <- valid %>%
  filter(
    complete.cases(
      class, MP, BPM, USG, ORTG,
      team_wins, barthag, `Conf Win%`, Role
    )
  )

test_model <- test %>%
  filter(
    complete.cases(
      class, MP, BPM, USG, ORTG,
      team_wins, barthag, `Conf Win%`, Role
    )
  )

glm_fit <- glm(
  transfer_next_year ~ GP + ORTG + USG + ORB. + TP. + major*PORPAG + adjoe*adjde + `Conf Win%` + Role + class + MP*BPM + major*BPM + player_team_gap*bpm_minutes_gap + mid_major_star*underused_star,
  data = train_model,
  family = binomial()
)

valid_model$pred_glm <- predict(glm_fit, newdata = valid_model, type = "response")
test_model$pred_glm  <- predict(glm_fit, newdata = test_model, type = "response")

#eval

#tidy(glm_fit, exponentiate = TRUE, conf.int = TRUE)

auc_valid <- roc(valid_model$transfer_next_year, valid_model$pred_glm)
auc_valid
plot(auc_valid)

valid_model <- valid_model %>%
  mutate(pred_class = ifelse(pred_glm > 0.33, 1, 0))
table(
  Actual = valid_model$transfer_next_year,
  Predicted = valid_model$pred_class
)
confusionMatrix(
  factor(valid_model$pred_class),
  factor(valid_model$transfer_next_year)
)

valid_model <- valid_model %>%
  mutate(prob_bin = ntile(pred_glm, 10))
valid_model %>%
  group_by(prob_bin) %>%
  summarise(
    avg_pred = mean(pred_glm),
    actual_rate = mean(transfer_next_year),
    n = n()
  )

vif(glm_fit)

tidy(glm_fit) %>%
  filter(term != "(Intercept)") %>%
  mutate(abs_coef = abs(estimate)) %>%
  arrange(desc(abs_coef))

valid_model %>%
  arrange(desc(pred_glm)) %>%
  select(player, team, class, MP, BPM, USG, team_wins, pred_glm) %>%
  head(25)

valid_model %>%
  filter(transfer_next_year == 1) %>%
  arrange(pred_glm) %>%
  select(player, team, class, MP, BPM, USG, team_wins, pred_glm) %>%
  head(25)

ggplot(valid_model, aes(pred_glm, fill = factor(transfer_next_year))) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  labs(fill = "Transfer")

test_model$pred_class <- ifelse(test_model$pred_glm > 0.33, 1, 0)
confusionMatrix(
  factor(test_model$pred_class),
  factor(test_model$transfer_next_year)
)
roc(test_model$transfer_next_year, test_model$pred_glm)

#confidence

boot_pred <- function(data, indices) {
  d <- data[indices,]

  fit <- glm(
    transfer_next_year ~ GP + ORTG + USG + ORB. + TP. +
      major*PORPAG + adjoe*adjde + `Conf Win%` +
      Role + class + MP*BPM +
      major*BPM +
      player_team_gap*bpm_minutes_gap +
      mid_major_star*underused_star,
    data = d,
    family = binomial()
  )

  predict(fit, newdata = valid_model, type = "response")
}

boot_results <- boot(
  data = train_model,
  statistic = boot_pred,
  R = 200
)

pred_matrix <- boot_results$t

valid_model <- valid_model %>%
  mutate(
    prob_mean = colMeans(pred_matrix),
    lower_ci = apply(pred_matrix, 2, quantile, 0.025),
    upper_ci = apply(pred_matrix, 2, quantile, 0.975)
  )

valid_model <- valid_model %>%
  mutate(
    uncertainty = upper_ci - lower_ci,
    confidence_score = 1 - uncertainty
  ) %>%
  mutate(
    confidence_score = scales::rescale(confidence_score, to=c(0,1))
  )

boot_pred_test <- function(data, indices) {
  d <- data[indices,]

  fit <- glm(
    transfer_next_year ~ GP + ORTG + USG + ORB. + TP. +
      major*PORPAG + adjoe*adjde + `Conf Win%` +
      Role + class + MP*BPM +
      major*BPM +
      player_team_gap*bpm_minutes_gap +
      mid_major_star*underused_star,
    data = d,
    family = binomial()
  )

  predict(fit, newdata = test_model, type = "response")
}

boot_results_test <- boot(
  data = train_model,
  statistic = boot_pred_test,
  R = 200
)

pred_matrix_test <- boot_results_test$t

test_model <- test_model %>%
  mutate(
    prob_mean = colMeans(pred_matrix_test),
    lower_ci = apply(pred_matrix_test, 2, quantile, 0.025),
    upper_ci = apply(pred_matrix_test, 2, quantile, 0.975)
  )

test_model <- test_model %>%
  mutate(
    uncertainty = upper_ci - lower_ci,
    confidence_score = 1 - uncertainty
  ) %>%
  mutate(
    confidence_score = scales::rescale(confidence_score, to = c(0,1))
  )

transfer_formula <- formula(glm_fit)
```

return vs leave

```{r, eval=FALSE}
return_leave_base <- model_table_START %>%
  mutate(
    year = as.integer(year),
    pid = as.character(pid),
    player = str_squish(player),
    team = str_squish(team),

    conf = str_squish(conf),
    conf = str_to_lower(conf),
    conf = if_else(conf == "ind", "independent", conf),

    major = case_when(
      conf %in% c("b12", "acc", "b10", "sec", "be", "p12") ~ "high",
      TRUE ~ "mid"
    ),

    class = str_squish(class),
    class = na_if(class, ""),
    class = na_if(class, "N/A"),
    class = na_if(class, "--"),
    class = case_when(
      class %in% c("Fr", "So", "Jr", "Sr") ~ class,
      TRUE ~ NA_character_
    ),

    Role = str_squish(Role),
    Role = na_if(Role, ""),

    Birthdate = as.character(Birthdate),
    Birthdate = na_if(Birthdate, ""),
    birth_date = suppressWarnings(mdy(Birthdate)),
    age = if_else(
      !is.na(birth_date),
      as.numeric(interval(birth_date, ymd(paste0(year, "-11-01"))) / years(1)),
      NA_real_
    ),

    class_num = case_when(
      class == "Fr" ~ 1,
      class == "So" ~ 2,
      class == "Jr" ~ 3,
      class == "Sr" ~ 4,
      TRUE ~ NA_real_
    ),

    underclass = case_when(
      class %in% c("Fr", "So") ~ "Underclassmen",
      class %in% c("Jr", "Sr") ~ "Upperclassmen",
      TRUE ~ NA_character_
    ),

    player_team_gap = BPM - barthag
  ) %>%
  group_by(team, year) %>%
  mutate(
    team_total_mp = sum(MP, na.rm = TRUE),
    team_total_gp = sum(GP, na.rm = TRUE),

    minutes_share = if_else(team_total_mp > 0, MP / team_total_mp, NA_real_),
    games_share = if_else(team_total_gp > 0, GP / team_total_gp, NA_real_),

    mp_rank_team = min_rank(desc(MP)),
    bpm_rank_team = min_rank(desc(BPM)),
    usg_rank_team = min_rank(desc(USG)),
    porpag_rank_team = min_rank(desc(PORPAG)),

    mp_pct_team = percent_rank(MP),
    bpm_pct_team = percent_rank(BPM),
    usg_pct_team = percent_rank(USG),
    porpag_pct_team = percent_rank(PORPAG)
  ) %>%
  ungroup() %>%
  group_by(conf, year) %>%
  mutate(
    bpm_pct_conf = percent_rank(BPM),
    porpag_pct_conf = percent_rank(PORPAG),
    mp_pct_conf = percent_rank(MP),
    usg_pct_conf = percent_rank(USG)
  ) %>%
  ungroup() %>%
  mutate(
    bpm_minutes_gap = BPM - minutes_share,

    underused_good_player = if_else(BPM > 4 & mp_rank_team > 5, 1L, 0L),
    low_role_high_output = if_else(porpag_pct_conf >= 0.75 & mp_pct_team <= 0.40, 1L, 0L),
    mid_major_star = if_else(major == "mid" & bpm_pct_conf >= 0.90, 1L, 0L),
    high_major_low_role = if_else(major == "high" & mp_pct_team <= 0.25, 1L, 0L),

    winning_team = if_else(!is.na(team_wins) & !is.na(team_losses) & team_wins > team_losses, 1L, 0L),
    conf_winning_team = if_else(
      !is.na(conf_wins) & !is.na(conf_losses) & conf_wins > conf_losses,
      1L, 0L
    )
  ) %>%
  mutate(
    conf = factor(conf),
    major = factor(major, levels = c("high", "mid")),
    class = factor(class, levels = c("Fr", "So", "Jr", "Sr")),
    underclass = factor(underclass, levels = c("Underclassmen", "Upperclassmen")),
    Role = factor(Role)
  )

next_year_lookup <- return_leave_base %>%
  select(
    pid, year, player, Birthdate,
    next_team = team,
    next_conf = conf,
    next_class = class,
    next_role = Role
  ) %>%
  distinct() %>%
  mutate(year = year - 1)

return_leave_table <- return_leave_base %>%
  left_join(
    next_year_lookup,
    by = c("pid", "year", "player", "Birthdate")
  ) %>%
  mutate(
    returned_next_year = if_else(!is.na(next_team), 1L, 0L),
    left_next_year = if_else(is.na(next_team), 1L, 0L),

    return_leave_status = case_when(
      returned_next_year == 1L ~ "returned",
      returned_next_year == 0L ~ "left",
      TRUE ~ NA_character_
    ),

    same_team_next_year = case_when(
      is.na(next_team) ~ NA_integer_,
      next_team == team ~ 1L,
      next_team != team ~ 0L
    ),

    transferred_if_returned = case_when(
      is.na(next_team) ~ NA_integer_,
      next_team != team ~ 1L,
      next_team == team ~ 0L
    )
  ) %>%
  mutate(
    return_leave_status = factor(return_leave_status, levels = c("returned", "left"))
  )

last_observed_year <- max(return_leave_table$year, na.rm = TRUE)

model_data_return_leave <- return_leave_table %>%
  filter(year < last_observed_year)

train_return <- model_data_return_leave %>% filter(year <= 2022)
valid_return <- model_data_return_leave %>% filter(year == 2023)
test_return <- model_data_return_leave %>% filter(year == 2024)

class_levels <- levels(droplevels(train_return$class))
role_levels  <- levels(droplevels(train_return$Role))
conf_levels  <- levels(droplevels(train_return$conf))
major_levels <- levels(droplevels(train_return$major))
underclass_levels <- levels(droplevels(train_return$underclass))

train_return <- train_return %>%
  mutate(
    class = factor(class, levels = class_levels),
    Role = factor(Role, levels = role_levels),
    conf = factor(conf, levels = conf_levels),
    major = factor(major, levels = major_levels),
    underclass = factor(underclass, levels = underclass_levels)
  )

valid_return <- valid_return %>%
  mutate(
    class = factor(class, levels = class_levels),
    Role = factor(Role, levels = role_levels),
    conf = factor(conf, levels = conf_levels),
    major = factor(major, levels = major_levels),
    underclass = factor(underclass, levels = underclass_levels)
  )

test_return <- test_return %>%
  mutate(
    class = factor(class, levels = class_levels),
    Role = factor(Role, levels = role_levels),
    conf = factor(conf, levels = conf_levels),
    major = factor(major, levels = major_levels),
    underclass = factor(underclass, levels = underclass_levels)
  )

glm_return <- glm(
  returned_next_year ~ GP + ORTG + USG + ORB. + major*PORPAG + adjoe*adjde + `Conf Win%` + Role + class + MP*BPM + major*BPM + bpm_minutes_gap + mid_major_star*underused_good_player,
  data = train_return,
  family = binomial()
)

summary(glm_return)

valid_return$pred_glm <- predict(glm_return, newdata = valid_return, type = "response")
test_return$pred_glm  <- predict(glm_return, newdata = test_return, type = "response")

auc_valid_return <- roc(valid_return$returned_next_year, valid_return$pred_glm)
auc_valid_return
plot(auc_valid_return)

auc_test_return <- roc(test_return$returned_next_year, test_return$pred_glm)
auc_test_return
plot(auc_test_return)

valid_model_return <- valid_return %>%
  mutate(pred_class = ifelse(pred_glm > 0.50, 1, 0))

table(
  Actual = valid_model_return$returned_next_year,
  Predicted = valid_model_return$pred_class
)

confusionMatrix(
  factor(valid_model_return$pred_class),
  factor(valid_model_return$returned_next_year)
)

test_model_return <- test_return %>%
  mutate(pred_class = ifelse(pred_glm > 0.50, 1, 0))

table(
  Actual = test_model_return$returned_next_year,
  Predicted = test_model_return$pred_class
)

confusionMatrix(
  factor(test_model_return$pred_class),
  factor(test_model_return$returned_next_year)
)

valid_model_return %>%
  mutate(prob_bin = ntile(pred_glm, 10)) %>%
  group_by(prob_bin) %>%
  summarise(
    avg_pred = mean(pred_glm),
    actual_rate = mean(returned_next_year),
    n = n(),
    .groups = "drop"
  )

test_model_return %>%
  mutate(prob_bin = ntile(pred_glm, 10)) %>%
  group_by(prob_bin) %>%
  summarise(
    avg_pred = mean(pred_glm),
    actual_rate = mean(returned_next_year),
    n = n(),
    .groups = "drop"
  )

# tidy(glm_return, exponentiate = TRUE, conf.int = TRUE) %>%
#   arrange(p.value)
# 
# tidy(glm_return) %>%
#   filter(term != "(Intercept)") %>%
#   mutate(abs_coef = abs(estimate)) %>%
#   arrange(desc(abs_coef))

vif(glm_return)

auc_valid_return
auc_test_return

#confidence

boot_pred_return_valid <- function(data, indices) {
  d <- data[indices, ]

  fit <- glm(
    returned_next_year ~ GP + ORTG + USG + ORB. + TP. +
      major*PORPAG + adjoe*adjde + `Conf Win%` +
      Role + class + MP*BPM +
      major*BPM +
      bpm_minutes_gap +
      mid_major_star*underused_good_player,
    data = d,
    family = binomial()
  )

  predict(fit, newdata = valid_model_return, type = "response")
}
set.seed(123)

boot_results_valid_return <- boot(
  data = train_return,
  statistic = boot_pred_return_valid,
  R = 200
)

pred_matrix_valid_return <- boot_results_valid_return$t

valid_model_return <- valid_model_return %>%
  mutate(
    prob_mean = colMeans(pred_matrix_valid_return, na.rm = TRUE),
    lower_ci = apply(pred_matrix_valid_return, 2, quantile, probs = 0.025, na.rm = TRUE),
    upper_ci = apply(pred_matrix_valid_return, 2, quantile, probs = 0.975, na.rm = TRUE)
  ) %>%
  mutate(
    uncertainty = upper_ci - lower_ci,
    confidence_score = 1 - uncertainty
  ) %>%
  mutate(
    confidence_score = rescale(confidence_score, to = c(0, 1))
  )

boot_pred_return_test <- function(data, indices) {
  d <- data[indices, ]

  fit <- glm(
    returned_next_year ~ GP + ORTG + USG + ORB. + TP. +
      major*PORPAG + adjoe*adjde + `Conf Win%` +
      Role + class + MP*BPM +
      major*BPM +
      bpm_minutes_gap +
      mid_major_star*underused_good_player,
    data = d,
    family = binomial()
  )

  predict(fit, newdata = test_model_return, type = "response")
}
set.seed(123)

boot_results_test_return <- boot(
  data = train_return,
  statistic = boot_pred_return_test,
  R = 200
)

pred_matrix_test_return <- boot_results_test_return$t

test_model_return <- test_model_return %>%
  mutate(
    prob_mean = colMeans(pred_matrix_test_return, na.rm = TRUE),
    lower_ci = apply(pred_matrix_test_return, 2, quantile, probs = 0.025, na.rm = TRUE),
    upper_ci = apply(pred_matrix_test_return, 2, quantile, probs = 0.975, na.rm = TRUE)
  ) %>%
  mutate(
    uncertainty = upper_ci - lower_ci,
    confidence_score = 1 - uncertainty
  ) %>%
  mutate(
    confidence_score = rescale(confidence_score, to = c(0, 1))
  )

return_leave_formula <- formula(glm_return)
```

destination conference model

```{r, eval=FALSE}
model_table <- model_table_START %>%
  mutate(
    year = as.integer(year),
    pid = as.character(pid),
    player = str_squish(player),
    team = str_squish(team),

    conf = str_squish(conf),
    conf = str_to_lower(conf),
    conf = if_else(conf == "ind", "independent", conf),

    class = str_squish(class),
    class = na_if(class, ""),
    class = na_if(class, "N/A"),
    class = na_if(class, "--"),
    class = case_when(
      class %in% c("Fr", "So", "Jr", "Sr") ~ class,
      TRUE ~ NA_character_
    ),

    Role = str_squish(Role),
    Role = na_if(Role, "")
  )

high_major_confs <- c("acc", "b10", "b12", "sec", "be", "p12") #6 really 5
upper_mid_confs  <- c("amer", "a10", "wcc", "mwc", "mvc", 'wac', 'caa', 'ivy', 'mac', 'cusa') #10
low_major_confs  <- c("ae", "asun", "bsth", "bsky", "bw", "horz", "maac", "meac", "nec", "ovc", "pat", "sb", "sc", "slnd", "sum", "swac", "ind", "independent") #18

get_conf_tier <- function(conf) {
  case_when(
    conf %in% high_major_confs ~ "high_major",
    conf %in% upper_mid_confs  ~ "mid_major",
    conf %in% low_major_confs  ~ "low_major",
    is.na(conf)                ~ NA_character_
  )
}

model_table <- model_table %>%
  mutate(
    conf_tier = get_conf_tier(conf),
    conf_tier = factor(conf_tier,
                       levels = c("high_major", "mid_major", "low_major"))
  )

next_year_lookup <- model_table %>%
  transmute(
    pid,
    year = year - 1,
    next_team = team,
    next_conf = conf,
    next_conf_tier = conf_tier
  )

model_table <- model_table %>%
  left_join(next_year_lookup, by = c("pid", "year"))

model_table <- model_table %>%
  mutate(
    has_next_year = !is.na(next_team),
    transfer_next_year = if_else(has_next_year & next_team != team, 1, 0, missing = 0),
    stayed_next_year   = if_else(has_next_year & next_team == team, 1, 0, missing = 0),

    destination_conf = if_else(transfer_next_year == 1, next_conf, NA_character_),
    destination_tier = if_else(transfer_next_year == 1,
                               as.character(next_conf_tier),
                               NA_character_),

    same_conf_transfer = if_else(
      transfer_next_year == 1 & conf == next_conf, 1, 0, missing = 0
    )
  )

tier_num_map <- c(
  "low_major" = 1,
  "mid_major" = 2,
  "high_major" = 3
)

model_table <- model_table %>%
  mutate(
    current_tier_num = tier_num_map[as.character(conf_tier)],
    dest_tier_num    = tier_num_map[as.character(destination_tier)],
    movement_type = case_when(
      transfer_next_year == 1 & dest_tier_num > current_tier_num ~ "up",
      transfer_next_year == 1 & dest_tier_num < current_tier_num ~ "down",
      transfer_next_year == 1 & dest_tier_num == current_tier_num ~ "lateral",
      TRUE ~ NA_character_
    )
  )

model_table <- model_table %>%
  group_by(team, year) %>%
  mutate(
    team_total_mp = sum(MP, na.rm = TRUE),
    minutes_share = if_else(team_total_mp > 0, MP / team_total_mp, NA_real_),

    mp_rank_team = min_rank(desc(MP)),
    bpm_rank_team = min_rank(desc(BPM)),
    usg_rank_team = min_rank(desc(USG)),
    porpag_rank_team = min_rank(desc(PORPAG)),

    bpm_pct_team = percent_rank(BPM),
    mp_pct_team = percent_rank(MP),
    usg_pct_team = percent_rank(USG),
    porpag_pct_team = percent_rank(PORPAG),

    bpm_minutes_gap = bpm_pct_team - mp_pct_team
  ) %>%
  ungroup() %>%
  group_by(conf, year) %>%
  mutate(
    bpm_pct_conf = percent_rank(BPM),
    porpag_pct_conf = percent_rank(PORPAG),
    mp_pct_conf = percent_rank(MP)
  ) %>%
  ungroup() %>%
  mutate(
    underclass = if_else(class %in% c("Fr", "So"), 1, 0, missing = 0),
    mid_major_star = if_else(conf_tier %in% c("mid_major", "low_major") &
                               BPM >= 8 & PORPAG >= 2.5, 1, 0, missing = 0),
    underused_star = if_else(BPM >= 6 & mp_rank_team > 5, 1, 0, missing = 0),
    high_major_role_player = if_else(conf_tier == "high_major" &
                                       MP >= 10 & MP <= 25 &
                                       BPM >= 0, 1, 0, missing = 0)
  )

destination_model_table <- model_table %>%
  filter(transfer_next_year == 1) %>%
  filter(!is.na(destination_conf), !is.na(destination_tier)) %>%
  filter(!is.na(BPM), !is.na(MP), !is.na(PORPAG), !is.na(USG), !is.na(ORTG)) %>%
  mutate(
    destination_tier = factor(destination_tier,
                              levels = c("high_major", "mid_major", "low_major")),
    class = factor(class, levels = c("Fr", "So", "Jr", "Sr")),
    Role = fct_lump_min(as.factor(Role), min = 25),
    conf_tier = factor(conf_tier,
                       levels = c("high_major", "mid_major", "low_major"))
  )

train_dest <- destination_model_table %>% filter(year <= 2022)
valid_dest <- destination_model_table %>% filter(year == 2023)
test_dest  <- destination_model_table %>% filter(year == 2024)

dest_tier_model <- multinom(
  destination_tier ~ BPM + MP + PORPAG + USG + ORTG +
    conf_tier + class + Role +
    minutes_share + bpm_minutes_gap +
    bpm_pct_team + porpag_pct_team + bpm_pct_conf +
    team_wins + `Conf Win%` + barthag +
    mid_major_star + underused_star + high_major_role_player,
  data = train_dest,
  trace = FALSE
)

summary(dest_tier_model)

valid_tier_probs <- predict(dest_tier_model, newdata = valid_dest, type = "probs")
test_tier_probs  <- predict(dest_tier_model, newdata = test_dest, type = "probs")

valid_pred_class <- colnames(valid_tier_probs)[max.col(valid_tier_probs, ties.method = "first")]
test_pred_class  <- colnames(test_tier_probs)[max.col(test_tier_probs, ties.method = "first")]

valid_results <- valid_dest %>%
  bind_cols(as.data.frame(valid_tier_probs)) %>%
  mutate(pred_destination_tier = valid_pred_class)

test_results <- test_dest %>%
  bind_cols(as.data.frame(test_tier_probs)) %>%
  mutate(pred_destination_tier = test_pred_class)

confusionMatrix(
  data = factor(valid_results$pred_destination_tier,
                levels = levels(valid_results$destination_tier)),
  reference = valid_results$destination_tier
)

confusionMatrix(
  data = factor(test_results$pred_destination_tier,
                levels = levels(test_results$destination_tier)),
  reference = test_results$destination_tier
)

top_k_accuracy <- function(probs, actual, k = 3) {
  probs <- as.data.frame(probs)

  topk <- apply(probs, 1, function(x) {
    names(sort(x, decreasing = TRUE))[1:k]
  })

  # transpose so each row = one observation
  topk <- t(topk)

  mean(sapply(1:nrow(topk), function(i) {
    actual[i] %in% topk[i, ]
  }))
}

log_loss <- function(actual, probs) {
  probs <- as.matrix(probs)
  actual <- as.factor(actual)

  # create one-hot encoding
  actual_mat <- model.matrix(~ actual - 1)

  # align column names
  colnames(actual_mat) <- levels(actual)

  # reorder probs to match
  probs <- probs[, colnames(actual_mat)]

  eps <- 1e-15
  probs <- pmax(pmin(probs, 1 - eps), eps)

  -mean(rowSums(actual_mat * log(probs)))
}

top3_acc <- top_k_accuracy(
  probs = valid_tier_probs,
  actual = valid_dest$destination_tier,
  k = 3
)
top3_acc

ll <- log_loss(
  actual = valid_dest$destination_tier,
  probs = valid_tier_probs
)
ll

top3_test <- top_k_accuracy(
  probs = test_tier_probs,
  actual = test_dest$destination_tier,
  k = 3
)
top3_test

ll_test <- log_loss(
  actual = test_dest$destination_tier,
  probs = test_tier_probs
)
ll_test

destination_tier_formula <- formula(dest_tier_model)
```

#projected stats

```{r, eval=FALSE}
high_major_confs <- c("acc", "b10", "b12", "sec", "be", "p12")
upper_mid_confs  <- c("amer", "a10", "wcc", "mwc", "mvc", "wac", "caa", "ivy", "mac", "cusa")
low_major_confs  <- c("ae", "asun", "bsth", "bsky", "bw", "horz", "maac", "meac", "nec", "ovc", 
                      "pat", "sb", "sc", "slnd", "sum", "swac", "ind", "independent")

mt <- model_table_START %>%
  mutate(
    conf  = tolower(trimws(conf)),
    team  = trimws(team),
    player = trimws(player),
    class = trimws(class)
  )

next_year_lookup <- mt %>%
  select(
    pid,
    year,
    team_next   = team,
    conf_next   = conf,
    BPM_next    = BPM,
    PORPAG_next = PORPAG,
    MP_next     = MP,
    USG_next    = USG,
    team_wins_next = team_wins,
    barthag_next   = barthag
  ) %>%
  mutate(year = year - 1)   # so year t row joins to year t+1 info

mt2 <- mt %>%
  left_join(next_year_lookup, by = c("pid", "year"))

mt2 <- mt2 %>%
  mutate(
    transfer = if_else(!is.na(team_next) & team != team_next, 1, 0),
    
    conf_tier_next = case_when(
      conf_next %in% high_major_confs ~ "high",
      conf_next %in% upper_mid_confs  ~ "upper_mid",
      conf_next %in% low_major_confs  ~ "low",
      TRUE ~ NA_character_
    ),
    
    conf_tier_current = case_when(
      conf %in% high_major_confs ~ "high",
      conf %in% upper_mid_confs  ~ "upper_mid",
      conf %in% low_major_confs  ~ "low",
      TRUE ~ NA_character_
    ),
    
    conf_tier_next = factor(conf_tier_next, levels = c("low", "upper_mid", "high")),
    conf_tier_current = factor(conf_tier_current, levels = c("low", "upper_mid", "high")),
    
    underclass = if_else(class %in% c("Fr", "So"), 1, 0)
  )

model_data <- mt2 %>%
  filter(transfer == 1) %>%
  filter(!is.na(BPM_next)) %>%
  filter(!is.na(conf_tier_next)) %>%
  filter(!is.na(BPM), !is.na(PORPAG), !is.na(MP), !is.na(USG),
         !is.na(adjoe), !is.na(adjde), !is.na(barthag)) %>%
  filter(class %in% c("Fr", "So", "Jr", "Sr"))

model_data <- model_data %>%
  group_by(team, year) %>%
  mutate(
    minutes_share = MP / sum(MP, na.rm = TRUE),
    mp_rank_team  = min_rank(desc(MP)),
    bpm_rank_team = min_rank(desc(BPM)),
    porpag_rank_team = min_rank(desc(PORPAG))
  ) %>%
  ungroup()

train <- model_data %>% filter(year <= 2022)
valid <- model_data %>% filter(year == 2023)
test  <- model_data %>% filter(year == 2024)

# quick size check
cat("Train n:", nrow(train), "\n")
cat("Valid n:", nrow(valid), "\n")
cat("Test n :", nrow(test), "\n")

form <- BPM_next ~ 
  BPM * conf_tier_next +
  PORPAG * conf_tier_next +
  minutes_share +
  MP +
  USG +
  adjoe +
  adjde +
  barthag +
  mp_rank_team +
  bpm_rank_team +
  porpag_rank_team +
  underclass +
  class +
  conf_tier_current

x_train <- model.matrix(form, data = train)[, -1]
y_train <- train$BPM_next

x_valid <- model.matrix(form, data = valid)[, -1]
y_valid <- valid$BPM_next

x_test  <- model.matrix(form, data = test)[, -1]
y_test  <- test$BPM_next

set.seed(123)

cv_fit <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 0,          # ridge
  family = "gaussian",
  nfolds = 5
)

best_lambda <- cv_fit$lambda.min
best_lambda

valid_pred <- as.numeric(predict(cv_fit, s = best_lambda, newx = x_valid))
test_pred  <- as.numeric(predict(cv_fit, s = best_lambda, newx = x_test))

rmse_fun <- function(actual, pred) {
  sqrt(mean((actual - pred)^2, na.rm = TRUE))
}

mae_fun <- function(actual, pred) {
  mean(abs(actual - pred), na.rm = TRUE)
}

r2_fun <- function(actual, pred) {
  ss_res <- sum((actual - pred)^2, na.rm = TRUE)
  ss_tot <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  1 - ss_res / ss_tot
}

eval_metrics <- function(actual, pred) {
  tibble(
    RMSE = rmse_fun(actual, pred),
    MAE  = mae_fun(actual, pred),
    R2   = r2_fun(actual, pred)
  )
}

# overall
valid_results <- eval_metrics(y_valid, valid_pred)
test_results  <- eval_metrics(y_test, test_pred)

valid_results
test_results

valid_out <- valid %>%
  mutate(pred_BPM_next = valid_pred)

test_out <- test %>%
  mutate(pred_BPM_next = test_pred)

valid_by_tier <- valid_out %>%
  group_by(conf_tier_next) %>%
  summarise(
    n = n(),
    RMSE = rmse_fun(BPM_next, pred_BPM_next),
    MAE  = mae_fun(BPM_next, pred_BPM_next),
    R2   = r2_fun(BPM_next, pred_BPM_next),
    .groups = "drop"
  )

test_by_tier <- test_out %>%
  group_by(conf_tier_next) %>%
  summarise(
    n = n(),
    RMSE = rmse_fun(BPM_next, pred_BPM_next),
    MAE  = mae_fun(BPM_next, pred_BPM_next),
    R2   = r2_fun(BPM_next, pred_BPM_next),
    .groups = "drop"
  )

valid_by_tier
test_by_tier

coef_table <- coef(cv_fit, s = "lambda.min") %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("term")

# rename the ONLY non-term column automatically
names(coef_table)[2] <- "coef"

coef_table <- coef_table %>%
  arrange(desc(abs(coef)))

head(coef_table, 25)

test_predictions_table <- test_out %>%
  select(player, team, conf, team_next, conf_next, conf_tier_next, BPM, BPM_next, pred_BPM_next) %>%
  mutate(error = BPM_next - pred_BPM_next) %>%
  arrange(desc(pred_BPM_next))

head(test_predictions_table, 25)

bpm_formula <- formula(form)
```

save training metadata
```{r}
transfer_meta <- list(
  formula = transfer_formula,
  class_levels = levels(train_model$class),
  role_levels = levels(train_model$Role),
  threshold = 0.33
)

return_meta <- list(
  formula = return_leave_formula,
  class_levels = levels(train_return$class),
  role_levels = levels(train_return$Role),
  threshold = 0.50
)

dest_meta <- list(
  formula = formula(dest_tier_model),
  class_levels = levels(train_dest$class),
  role_levels = levels(train_dest$Role),
  conf_tier_levels = levels(train_dest$conf_tier),
  destination_tier_levels = levels(train_dest$destination_tier)
)

bpm_meta <- list(
  formula = bpm_formula,
  lambda = best_lambda,
  class_levels = c("Fr", "So", "Jr", "Sr"),
  conf_tier_current_levels = c("low", "upper_mid", "high"),
  conf_tier_next_levels = c("low", "upper_mid", "high"),
  x_colnames = colnames(x_train)
)
```

master feature engineering function
```{r}
engineer_portal_features <- function(df) {
  high_major_confs <- c("b12", "acc", "b10", "sec", "be", "p12")
  upper_mid_confs  <- c("amer", "a10", "wcc", "mwc", "mvc", "wac", "caa", "ivy", "mac", "cusa")
  low_major_confs  <- c("ae", "asun", "bsth", "bsky", "bw", "horz", "maac", "meac", "nec", "ovc",
                        "pat", "sb", "sc", "slnd", "sum", "swac", "ind", "independent")

  out <- df %>%
    mutate(
      year = as.integer(year),
      pid = as.character(pid),
      player = str_squish(player),
      team = str_squish(team),

      conf = str_squish(conf),
      conf = str_to_lower(conf),
      conf = if_else(conf == "ind", "independent", conf),

      class = str_squish(class),
      class = na_if(class, ""),
      class = na_if(class, "N/A"),
      class = na_if(class, "--"),
      class = case_when(
        class %in% c("Fr", "So", "Jr", "Sr") ~ class,
        TRUE ~ NA_character_
      ),

      Role = str_squish(Role),
      Role = na_if(Role, ""),

      Birthdate = as.character(Birthdate),
      Birthdate = na_if(Birthdate, ""),
      birth_date = suppressWarnings(mdy(Birthdate)),
      age = if_else(
        !is.na(birth_date),
        as.numeric(interval(birth_date, ymd(paste0(year, "-11-01"))) / years(1)),
        NA_real_
      ),

      major = case_when(
        conf %in% high_major_confs ~ "high",
        TRUE ~ "mid"
      ),

      conf_tier = case_when(
        conf %in% high_major_confs ~ "high_major",
        conf %in% upper_mid_confs  ~ "mid_major",
        conf %in% low_major_confs  ~ "low_major",
        TRUE ~ NA_character_
      ),

      class_num = case_when(
        class == "Fr" ~ 1,
        class == "So" ~ 2,
        class == "Jr" ~ 3,
        class == "Sr" ~ 4,
        TRUE ~ NA_real_
      ),

      underclass = case_when(
        class %in% c("Fr", "So") ~ "Underclassmen",
        class %in% c("Jr", "Sr") ~ "Upperclassmen",
        TRUE ~ NA_character_
      ),

      player_team_gap = BPM - barthag
    ) %>%
    group_by(team, year) %>%
    mutate(
      team_total_mp = sum(MP, na.rm = TRUE),
      team_total_gp = sum(GP, na.rm = TRUE),

      minutes_share = if_else(team_total_mp > 0, MP / team_total_mp, NA_real_),
      games_share   = if_else(team_total_gp > 0, GP / team_total_gp, NA_real_),

      mp_rank_team = min_rank(desc(MP)),
      bpm_rank_team = min_rank(desc(BPM)),
      usg_rank_team = min_rank(desc(USG)),
      porpag_rank_team = min_rank(desc(PORPAG)),

      mp_rank = min_rank(desc(MP)),
      bpm_rank = min_rank(desc(BPM)),
      usage_rank = min_rank(desc(USG)),

      mp_pct_team = percent_rank(MP),
      bpm_pct_team = percent_rank(BPM),
      usg_pct_team = percent_rank(USG),
      porpag_pct_team = percent_rank(PORPAG)
    ) %>%
    ungroup() %>%
    group_by(conf, year) %>%
    mutate(
      bpm_pct_conf = percent_rank(BPM),
      porpag_pct_conf = percent_rank(PORPAG),
      mp_pct_conf = percent_rank(MP),
      usg_pct_conf = percent_rank(USG)
    ) %>%
    ungroup() %>%
    group_by(major, year) %>%
    mutate(
      usg_rank_c1 = min_rank(desc(USG)),
      porpag_rank_c1 = min_rank(desc(PORPAG)),
      bpm_rank_c1 = min_rank(desc(BPM))
    ) %>%
    ungroup() %>%
    mutate(
      bpm_minutes_gap = BPM - minutes_share,

      underused_star = if_else(BPM > 5 & mp_rank > 5, 1L, 0L),
      underused_good_player = if_else(BPM > 4 & mp_rank_team > 5, 1L, 0L),
      low_role_high_output = if_else(porpag_pct_conf >= 0.75 & mp_pct_team <= 0.40, 1L, 0L),
      mid_major_star = if_else(major == "mid" & bpm_pct_conf >= 0.90, 1L, 0L),
      high_major_low_role = if_else(major == "high" & mp_pct_team <= 0.25, 1L, 0L),
      high_major_role_player = if_else(conf_tier == "high_major" & MP >= 10 & MP <= 25 & BPM >= 0, 1L, 0L),

      winning_team = if_else(!is.na(team_wins) & !is.na(team_losses) & team_wins > team_losses, 1L, 0L),
      conf_winning_team = if_else(
        !is.na(conf_wins) & !is.na(conf_losses) & conf_wins > conf_losses, 1L, 0L
      )
    )

  out
}

attach_next_year_outcomes <- function(df) {
  next_year_lookup <- df %>%
    select(
      pid, year, player, Birthdate,
      next_team = team,
      next_conf = conf,
      next_class = class,
      next_role = Role,
      next_conf_tier = conf_tier,
      BPM_next = BPM,
      PORPAG_next = PORPAG,
      MP_next = MP,
      USG_next = USG,
      team_wins_next = team_wins,
      barthag_next = barthag
    ) %>%
    distinct() %>%
    mutate(year = year - 1)

  df %>%
    left_join(next_year_lookup, by = c("pid", "year", "player", "Birthdate")) %>%
    mutate(
      returned_next_year = if_else(!is.na(next_team), 1L, 0L),
      transfer_next_year = case_when(
        is.na(next_team) ~ NA_integer_,
        next_team != team ~ 1L,
        next_team == team ~ 0L
      ),
      left_next_year = if_else(is.na(next_team), 1L, 0L),
      destination_conf = if_else(transfer_next_year == 1, next_conf, NA_character_),
      destination_tier = if_else(transfer_next_year == 1, next_conf_tier, NA_character_)
    )
}

model_table_FE   <- engineer_portal_features(model_table_START)
complete_2026_FE <- engineer_portal_features(complete_data_2026)

historical_model_table <- model_table_FE %>%
  attach_next_year_outcomes()
```

factor level alignment
```{r}
align_transfer_newdata <- function(df, meta) {
  df %>%
    mutate(
      class = factor(class, levels = meta$class_levels),
      Role  = factor(Role,  levels = meta$role_levels)
    )
}

align_return_newdata <- function(df, meta) {
  df %>%
    mutate(
      class = factor(class, levels = meta$class_levels),
      Role  = factor(Role,  levels = meta$role_levels)
    )
}

align_dest_newdata <- function(df, meta) {
  df %>%
    mutate(
      class = factor(class, levels = meta$class_levels),
      Role = factor(Role, levels = meta$role_levels),
      conf_tier = factor(conf_tier, levels = meta$conf_tier_levels)
    )
}
```

score 2026 data with transfer model
```{r}
score_transfer_model <- function(newdata, fit, meta) {
  newdata2 <- align_transfer_newdata(newdata, meta)

  preds <- predict(fit, newdata = newdata2, type = "response")

  newdata2 %>%
    mutate(
      pred_transfer = preds,
      pred_transfer_flag = if_else(pred_transfer >= meta$threshold, 1L, 0L)
    )
}
```

bootstrap confidence intervals
```{r}
bootstrap_transfer_predictions <- function(train_data, newdata, R = 200, seed = 123) {
  set.seed(seed)

  boot_fun <- function(data, indices) {
    d <- data[indices, ]

    fit <- glm(
      transfer_next_year ~ GP + ORTG + USG + ORB. + TP. +
        major*PORPAG + adjoe*adjde + `Conf Win%` +
        Role + class + MP*BPM +
        major*BPM +
        player_team_gap*bpm_minutes_gap +
        mid_major_star*underused_star,
      data = d,
      family = binomial()
    )

    predict(fit, newdata = newdata, type = "response")
  }

  boot_out <- boot::boot(
    data = train_data,
    statistic = boot_fun,
    R = R
  )

  pred_mat <- boot_out$t

  tibble(
    transfer_prob_mean = colMeans(pred_mat, na.rm = TRUE),
    transfer_prob_lo   = apply(pred_mat, 2, quantile, probs = 0.025, na.rm = TRUE),
    transfer_prob_hi   = apply(pred_mat, 2, quantile, probs = 0.975, na.rm = TRUE)
  )
}

complete_2026_transfer_base <- complete_2026_FE %>%
  align_transfer_newdata(transfer_meta) %>%
  filter(
    complete.cases(
      class, MP, BPM, USG, ORTG,
      team_wins, barthag, `Conf Win%`, Role
    )
  )

complete_2026_transfer_preds <- score_transfer_model(
  newdata = complete_2026_transfer_base,
  fit = glm_fit,
  meta = transfer_meta
)

transfer_ci_2026 <- bootstrap_transfer_predictions(
  train_data = train_model,
  newdata = complete_2026_transfer_base,
  R = 200
)

complete_2026_transfer_preds <- bind_cols(
  complete_2026_transfer_preds,
  transfer_ci_2026
)
```

score return v leave
```{r}
score_return_model <- function(newdata, fit, meta) {
  newdata2 <- align_return_newdata(newdata, meta)

  preds <- predict(fit, newdata = newdata2, type = "response")

  newdata2 %>%
    mutate(
      pred_return = preds,
      pred_leave = 1 - preds,
      pred_return_flag = if_else(pred_return >= meta$threshold, 1L, 0L)
    )
}
```

score destination conference tier model
```{r}
score_destination_model <- function(newdata, fit, meta) {
  newdata2 <- align_dest_newdata(newdata, meta)

  probs <- predict(fit, newdata = newdata2, type = "probs")
  probs <- as.data.frame(probs)

  names(probs) <- paste0("dest_prob_", names(probs))

  pred_class <- colnames(probs)[max.col(probs, ties.method = "first")]

  bind_cols(newdata2, probs) %>%
    mutate(pred_destination_tier = str_remove(pred_class, "^dest_prob_"))
}
```

score bpm under possible destination tier
```{r}
make_bpm_design_matrix <- function(df, meta) {
  rhs_terms <- delete.response(terms(meta$formula))
  mm <- model.matrix(rhs_terms, data = df)

  if ("(Intercept)" %in% colnames(mm)) {
    mm <- mm[, colnames(mm) != "(Intercept)", drop = FALSE]
  }

  missing_cols <- setdiff(meta$x_colnames, colnames(mm))
  if (length(missing_cols) > 0) {
    zero_mat <- matrix(0, nrow = nrow(mm), ncol = length(missing_cols))
    colnames(zero_mat) <- missing_cols
    mm <- cbind(mm, zero_mat)
  }

  extra_cols <- setdiff(colnames(mm), meta$x_colnames)
  if (length(extra_cols) > 0) {
    mm <- mm[, !(colnames(mm) %in% extra_cols), drop = FALSE]
  }

  mm <- mm[, meta$x_colnames, drop = FALSE]
  mm
}

score_bpm_by_tier <- function(newdata, fit, meta, tier_value) {
  nd <- newdata %>%
    mutate(
      conf_tier_next = factor(tier_value, levels = meta$conf_tier_next_levels),
      conf_tier_current = factor(conf_tier_current, levels = meta$conf_tier_current_levels),
      class = factor(class, levels = meta$class_levels),
      underclass = if_else(class %in% c("Fr", "So"), 1, 0)
    )

  x_new <- make_bpm_design_matrix(nd, meta)

  as.numeric(predict(fit, s = meta$lambda, newx = x_new))
}

complete_2026_bpm_base <- complete_2026_FE %>%
  mutate(
    conf_tier_current = case_when(
      conf %in% c("acc", "b10", "b12", "sec", "be", "p12") ~ "high",
      conf %in% c("amer", "a10", "wcc", "mwc", "mvc", "wac", "caa", "ivy", "mac", "cusa") ~ "upper_mid",
      conf %in% c("ae", "asun", "bsth", "bsky", "bw", "horz", "maac", "meac", "nec", "ovc",
                  "pat", "sb", "sc", "slnd", "sum", "swac", "ind", "independent") ~ "low",
      TRUE ~ NA_character_
    ),
    conf_tier_current = factor(conf_tier_current, levels = c("low", "upper_mid", "high")),
    class = factor(class, levels = c("Fr", "So", "Jr", "Sr"))
  ) %>%
  filter(
    !is.na(BPM), !is.na(PORPAG), !is.na(MP), !is.na(USG),
    !is.na(adjoe), !is.na(adjde), !is.na(barthag),
    !is.na(conf_tier_current), !is.na(class)
  )

complete_2026_bpm_preds <- complete_2026_bpm_base %>%
  mutate(
    pred_BPM_low = score_bpm_by_tier(., cv_fit, bpm_meta, "low"),
    pred_BPM_upper_mid = score_bpm_by_tier(., cv_fit, bpm_meta, "upper_mid"),
    pred_BPM_high = score_bpm_by_tier(., cv_fit, bpm_meta, "high")
  )
```

bpm confidence intervals
```{r}
bootstrap_bpm_predictions <- function(train_data, newdata, meta, R = 200, seed = 123) {
  set.seed(seed)

  rhs_terms <- delete.response(terms(meta$formula))

  one_boot_scenario <- function(tier_value) {
    pred_mat <- replicate(
      R,
      simplify = "matrix",
      expr = {
        idx <- sample(seq_len(nrow(train_data)), replace = TRUE)
        d <- train_data[idx, ] %>%
          mutate(
            conf_tier_next = factor(conf_tier_next, levels = meta$conf_tier_next_levels),
            conf_tier_current = factor(conf_tier_current, levels = meta$conf_tier_current_levels),
            class = factor(class, levels = meta$class_levels)
          )

        x_boot <- model.matrix(rhs_terms, data = d)
        if ("(Intercept)" %in% colnames(x_boot)) {
          x_boot <- x_boot[, colnames(x_boot) != "(Intercept)", drop = FALSE]
        }

        y_boot <- d$BPM_next

        fit_boot <- glmnet(
          x = x_boot,
          y = y_boot,
          alpha = 0,
          lambda = meta$lambda,
          family = "gaussian"
        )

        nd <- newdata %>%
          mutate(
            conf_tier_next = factor(tier_value, levels = meta$conf_tier_next_levels),
            conf_tier_current = factor(conf_tier_current, levels = meta$conf_tier_current_levels),
            class = factor(class, levels = meta$class_levels),
            underclass = if_else(class %in% c("Fr", "So"), 1, 0)
          )

        x_new <- model.matrix(rhs_terms, data = nd)
        if ("(Intercept)" %in% colnames(x_new)) {
          x_new <- x_new[, colnames(x_new) != "(Intercept)", drop = FALSE]
        }

        missing_cols <- setdiff(colnames(x_boot), colnames(x_new))
        if (length(missing_cols) > 0) {
          zero_mat <- matrix(0, nrow = nrow(x_new), ncol = length(missing_cols))
          colnames(zero_mat) <- missing_cols
          x_new <- cbind(x_new, zero_mat)
        }

        extra_cols <- setdiff(colnames(x_new), colnames(x_boot))
        if (length(extra_cols) > 0) {
          x_new <- x_new[, !(colnames(x_new) %in% extra_cols), drop = FALSE]
        }

        x_new <- x_new[, colnames(x_boot), drop = FALSE]

        as.numeric(predict(fit_boot, newx = x_new))
      }
    )

    tibble(
      !!paste0("pred_BPM_", tier_value, "_lo") := apply(pred_mat, 1, quantile, probs = 0.025, na.rm = TRUE),
      !!paste0("pred_BPM_", tier_value, "_hi") := apply(pred_mat, 1, quantile, probs = 0.975, na.rm = TRUE)
    )
  }

  bind_cols(
    one_boot_scenario("low"),
    one_boot_scenario("upper_mid"),
    one_boot_scenario("high")
  )
}

bpm_ci_2026 <- bootstrap_bpm_predictions(
  train_data = train,
  newdata = complete_2026_bpm_base,
  meta = bpm_meta,
  R = 200,
  seed = 123
)

complete_2026_bpm_preds <- complete_2026_bpm_base %>%
  mutate(
    pred_BPM_low = score_bpm_by_tier(., cv_fit, bpm_meta, "low"),
    pred_BPM_upper_mid = score_bpm_by_tier(., cv_fit, bpm_meta, "upper_mid"),
    pred_BPM_high = score_bpm_by_tier(., cv_fit, bpm_meta, "high")
  ) %>%
  bind_cols(bpm_ci_2026)
```

re do final 2026 scoring table
```{r}
complete_2026_transfer_base <- complete_2026_FE %>%
  align_transfer_newdata(transfer_meta) %>%
  filter(
    complete.cases(
      class, MP, BPM, USG, ORTG,
      team_wins, barthag, `Conf Win%`, Role
    )
  )

complete_2026_transfer_preds <- score_transfer_model(
  newdata = complete_2026_transfer_base,
  fit = glm_fit,
  meta = transfer_meta
)

transfer_ci_2026 <- bootstrap_transfer_predictions(
  train_data = train_model,
  newdata = complete_2026_transfer_base,
  R = 200,
  seed = 123
)

complete_2026_transfer_preds <- complete_2026_transfer_preds %>%
  bind_cols(transfer_ci_2026) %>%
  select(
    pid, year,
    pred_transfer,
    pred_transfer_flag,
    transfer_prob_mean,
    transfer_prob_lo,
    transfer_prob_hi
  )

complete_2026_return_base <- complete_2026_FE %>%
  align_return_newdata(return_meta) %>%
  filter(
    complete.cases(
      class, MP, BPM, USG, ORTG,
      team_wins, barthag, `Conf Win%`, Role
    )
  )

complete_2026_return_preds <- score_return_model(
  newdata = complete_2026_return_base,
  fit = glm_return,
  meta = return_meta
) %>%
  select(
    pid, year,
    pred_return,
    pred_leave,
    pred_return_flag
  )

complete_2026_dest_base <- complete_2026_FE %>%
  align_dest_newdata(dest_meta) %>%
  filter(
    !is.na(BPM),
    !is.na(MP),
    !is.na(PORPAG),
    !is.na(USG),
    !is.na(ORTG),
    !is.na(conf_tier),
    !is.na(class),
    !is.na(Role),
    !is.na(minutes_share),
    !is.na(bpm_minutes_gap),
    !is.na(bpm_pct_team),
    !is.na(porpag_pct_team),
    !is.na(bpm_pct_conf),
    !is.na(team_wins),
    !is.na(`Conf Win%`),
    !is.na(barthag)
  )

complete_2026_dest_preds <- score_destination_model(
  newdata = complete_2026_dest_base,
  fit = dest_tier_model,
  meta = dest_meta
)

complete_2026_dest_preds <- complete_2026_dest_preds %>%
  rename_with(~ str_replace_all(., "high_major", "high"), starts_with("dest_prob_")) %>%
  rename_with(~ str_replace_all(., "mid_major", "upper_mid"), starts_with("dest_prob_")) %>%
  rename_with(~ str_replace_all(., "low_major", "low"), starts_with("dest_prob_")) %>%
  select(
    pid, year,
    pred_destination_tier,
    starts_with("dest_prob_")
  )

complete_2026_bpm_base <- complete_2026_FE %>%
  mutate(
    conf_tier_current = case_when(
      conf %in% c("acc", "b10", "b12", "sec", "be", "p12") ~ "high",
      conf %in% c("amer", "a10", "wcc", "mwc", "mvc", "wac", "caa", "ivy", "mac", "cusa") ~ "upper_mid",
      conf %in% c("ae", "asun", "bsth", "bsky", "bw", "horz", "maac", "meac", "nec", "ovc",
                  "pat", "sb", "sc", "slnd", "sum", "swac", "ind", "independent") ~ "low",
      TRUE ~ NA_character_
    ),
    conf_tier_current = factor(conf_tier_current, levels = bpm_meta$conf_tier_current_levels),
    class = factor(class, levels = bpm_meta$class_levels)
  ) %>%
  filter(
    !is.na(BPM),
    !is.na(PORPAG),
    !is.na(MP),
    !is.na(USG),
    !is.na(adjoe),
    !is.na(adjde),
    !is.na(barthag),
    !is.na(conf_tier_current),
    !is.na(class),
    !is.na(minutes_share),
    !is.na(mp_rank_team),
    !is.na(bpm_rank_team),
    !is.na(porpag_rank_team)
  )

bpm_ci_2026 <- bootstrap_bpm_predictions(
  train_data = train,
  newdata = complete_2026_bpm_base,
  meta = bpm_meta,
  R = 200,
  seed = 123
)

complete_2026_bpm_preds <- complete_2026_bpm_base %>%
  mutate(
    pred_BPM_low = score_bpm_by_tier(., cv_fit, bpm_meta, "low"),
    pred_BPM_upper_mid = score_bpm_by_tier(., cv_fit, bpm_meta, "upper_mid"),
    pred_BPM_high = score_bpm_by_tier(., cv_fit, bpm_meta, "high")
  ) %>%
  bind_cols(bpm_ci_2026) %>%
  select(
    pid, year,
    pred_BPM_low,
    pred_BPM_upper_mid,
    pred_BPM_high,
    pred_BPM_low_lo,
    pred_BPM_low_hi,
    pred_BPM_upper_mid_lo,
    pred_BPM_upper_mid_hi,
    pred_BPM_high_lo,
    pred_BPM_high_hi
  )

complete_2026_scored <- complete_2026_FE %>%
  left_join(complete_2026_transfer_preds, by = c("pid", "year")) %>%
  left_join(complete_2026_return_preds,   by = c("pid", "year")) %>%
  left_join(complete_2026_dest_preds,     by = c("pid", "year")) %>%
  left_join(complete_2026_bpm_preds,      by = c("pid", "year"))

#dir.create("docs/data", recursive = TRUE, showWarnings = FALSE)
#write_csv(complete_2026_scored, "docs/data/complete_2026_scored.csv")
```

```{r, eval=FALSE}
#yo <- read.csv('docs/data/complete_2026_scored.csv')

yo <- yo %>%
  mutate(conf = tolower(conf)) %>%  # make sure everything is lowercase first
  mutate(conf = case_when(
    conf == "b10" ~ "Big Ten",
    conf == "b12" ~ "Big 12",
    conf == "acc" ~ "ACC",
    conf == "sec" ~ "SEC",
    conf == "be"  ~ "Big East",
    conf == "p12" ~ "Pac-12",
    
    conf == "amer" ~ "American",
    conf == "a10"  ~ "A-10",
    conf == "mwc"  ~ "MWC",
    conf == "wcc"  ~ "WCC",
    conf == "mvc"  ~ "MVC",
    conf == "wac"  ~ "WAC",
    conf == "caa"  ~ "CAA",
    conf == "ivy"  ~ "Ivy",
    conf == "mac"  ~ "MAC",
    conf == "cusa" ~ "C-USA",
    
    conf == "ae"   ~ "AE",
    conf == "asun" ~ "ASUN",
    conf == "bsth" ~ "Big South",
    conf == "bsky" ~ "Big Sky",
    conf == "bw"   ~ "Big West",
    conf == "horz" ~ "Horizon",
    conf == "maac" ~ "MAAC",
    conf == "meac" ~ "MEAC",
    conf == "nec"  ~ "NEC",
    conf == "ovc"  ~ "OVC",
    conf == "pat"  ~ "Patriot",
    conf == "sb"   ~ "Sun Belt",
    conf == "sc"   ~ "SoCon",
    conf == "slnd" ~ "Southland",
    conf == "sum"  ~ "Summit",
    conf == "swac" ~ "SWAC",
    
    conf %in% c("ind", "independent") ~ "Independent",
    
    TRUE ~ str_to_title(conf)  # fallback: just clean capitalization
  ))

yo <- yo %>%
  mutate(
    PTS = round(PTS,1),
    REB = round(REB,1),
    AST = round(AST,1)
  )
#write_csv(yo, "docs/data/complete_2026_scored.csv")
```

